OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-12 manuscript_nar.tex
Added manuscript in NAR template

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{pdflscape}

@@

\author[1,2,*]{Enrique Vidal}
\author[1,2]{Fran\c{c}ois le Dily}
\author[1,2]{Javier Quilez}
\author[1,2]{Ralph Stadhouders}
\author[1,2]{Yasmina Cuartero}
\author[1,2]{Thomas Graf}
\author[1,2,3,4]{Marc A.  Mart\'i-Renom}
\author[1,2]{Miguel Beato}
\author[1,2]{Guillaume J. Filion}
\affil[1]{Gene Regulation, Stem Cells and Cancer Program, Centre for
Genomic Regulation (CRG), The Barcelona Institute of Science and
Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain}
\affil[2]{Universitat Pompeu Fabra (UPF), Barcelona, Spain}
\affil[3]{CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028
Barcelona, Spain}
\affil[4]{ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}
\affil[*]{\textbf{Contact:} enrique.vidal@crg.eu}

@@

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 1 kbp.
\textit{OneD} is implemented as an R package
available at http://www.github.com/qenvio/dryhic.
\end{abstract}

%% --------------- Introduction --------------- %%

@@

@@

%\enlargethispage{12pt}

\section{Methods}

@@

\begin{landscape}

@@

\end{landscape}

@@

We developed the routines contained in the \texttt{dryhic} R package to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences at different
resolutions (not shown).

@@

%% --------------- Results --------------- %%

\section{Results}

@@

\begin{figure}
\centerline{\includegraphics[width=.99\textwidth]{img/figure1.pdf}}
\caption{Total amount of contacts per bin. A. Non linear relationship
between the number of restriction sites and the total number of
contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The brown line represents the raw
signal, the blue line represents the signal after bias correction, the
black line represents the signal after bias and copy number (CN) correction.
The long arm of chromosome 17 (the region corresponding to 20-80 Mbp) is
present in four copies, explaining that the signal is about twice higher
than for the short arm.}
\label{fig:totals}
\end{figure}

@@

\begin{figure}
	\centerline{\includegraphics[width=.99\textwidth]
{img/copy_number_figure2.pdf}}
\caption{Pearson correlation between the total number of contacts per bin
and an independent estimation of the copy number (COSMIC). Left panel
T47D breast cancer cell line, right panel K562 leukemia cell line.}
\label{fig:copy_number}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.99\textwidth]
  {img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotypes. A and
B. Average changes compared to the raw data on the T47D and K562 sets. The
bars represent 95\% confidence intervals centered on the mean difference
of the correlation score between a given correction method and the raw
data. C and D. ROC curves on the T47D and K562 sets.  The areas under the
curve are indicated in the bottom right corner. The color code is the same
as panels A and B. The brown lines correspond to raw matrices. All results
in this figure are based on Pearson correlations between the observed over
expected counts.}
\label{fig:aberrant}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.99\textwidth]
  {img/correlation_normal_figure4.pdf}}
\caption{
Results of the comparison between samples with normal karyotypes. A.
Average changes compared to the raw data on the mouse data set.  B. ROC
curves on the mouse data set. The legend is as in
Figure~\ref{fig:aberrant}.}
\label{fig:normal}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.99\textwidth]
  {img/figure_cnv_correction.pdf}}
\caption{
Copy number correction A. Detail of a Hi-C matrix normalized with
different methods. The central portion has an increased copy number, which
affects normalization. \textit{ICE} fades it away, \textit{LGF} enhances
it and \textit{OneD} reduces the signal by about half. B. Profile of the
total amount of contacts after copy numer correction. The plot shows the
same region as panel A. The brown line represents the raw signal, the blue
line represents the signal after bias correction, the black line
represents the signal after bias and copy number (CN) correction.}
\label{fig:cnv_correction}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.99\textwidth]
  {img/figure_benchmark_time.pdf}}
\caption{
Computing time of the bias correction methods. A. Total time for the
entire genome at 100 kbp resolution. Each dot corresponds to one sample.
The only method faster than ours under performs in all sample comparisons.
B. Time to correct a reduced genome (chr19-22) of one sample at different
resolutions. Note the logarithmic scale on the y-axis on both panels.}
\label{fig:times}
\end{figure}

%% --------------- Discussion --------------- %%

\section{Discussion and Conclusions}

@@

@@

\section*{Acknowledgements}

@@

\section*{Author contributions}

Conceptualization, E.V.; Methodology, E.V. and G.F.; Software, E.V. and
G.F.; Formal Analysis, E.V.; Investigation, E.V., F.D., Y.C. and R.S.;
Resources, M.B.; Data curation, J.Q., Writing – Original Draft, E.V. and
G.F., Writing – Review \& Editing, F.D. J.Q., R.S., Y.C, T.G., M.A.M-R.
and M.B.; Visualization, E.V.; Supervision, G.F. and M.B., Project
Administration, M.B.; Funding Acquisition, M.B., T.G., M.A.M-R. and G.F.

\section*{Funding}

@@

\vspace*{-12pt}

>>>>

\documentclass[a4,center,fleqn]{NAR}

% Enter dates of publication
\copyrightyear{2017}
\pubdate{** ***** ****}
\pubyear{2017}
\jvolume{**}
\jissue{**}

%\articlesubtype{This is the article type (optional)}

@@

\author{%
Enrique Vidal\,$^{1,2,*}$,
Fran\c{c}ois le Dily\,$^{1,2}$,
Javier Quilez\,$^{1,2}$,
Ralph Stadhouders\,$^{1,2}$,
Yasmina Cuartero\,$^{1,2,3}$,
Thomas Graf\,$^{1,2}$,
Marc A.  Mart\'i-Renom\,$^{1,2,3,4}$,
Miguel Beato\,$^{1,2}$,
and Guillaume J. Filion\,$^{1,2}$%
\footnote{To whom correspondence should be addressed.
Tel: +34 93 316 01 15; Fax: +34 93 316 00 99; Email: enrique.vidal@crg.eu}}

\address{%
$^{1}$Gene Regulation, Stem Cells and Cancer Program, Centre for
Genomic Regulation (CRG), The Barcelona Institute of Science and
Technology (BIST), Dr. Aiguader 88, 08003, Barcelona, Spain
and
$^{2}$Universitat Pompeu Fabra (UPF), Barcelona, Spain
and
$^{3}$CNAG-CRG, Centre for Genomic Regulation (CRG), Barcelona
Institute of Science and Technology (BIST), Baldiri i Reixac 4, 08028
Barcelona, Spain
and
$^{4}$ICREA, Pg. Lluís Companys 23, 08010 Barcelona, Spain}

% Affiliation must include:
% Department name, institution name, full road and district address,
% state, Zip or postal code, country

\history{%
Received January 1, ****;
Revised February 1, ****;
Accepted March 1, ****}

@@

\begin{abstract}
The three-dimensional conformation of genomes is an
essential component of their biological activity. The advent of the Hi-C
technology enabled an unprecedented progress in our understanding of
genome structures. However, Hi-C is subject to systematic biases that can
compromise downstream analyses. Several strategies have been proposed to
remove those biases, but the issue of abnormal karyotypes received little
attention. Many experiments are performed in cancer cell lines, which
typically harbor large-scale copy number variations that create visible
defects on the raw Hi-C maps. The consequences of these widespread
artifacts on the normalized maps are mostly unexplored.
We observed that current normalization methods are not robust to the
presence of large-scale copy number variations, potentially obscuring
biological differences and enhancing batch effects. To address this issue,
we developed an alternative approach designed to take into account
chromosomal abnormalities. The method, called \textit{OneD}, increases
reproducibility among replicates of Hi-C samples with abnormal karyotype,
outperforming previous methods significantly. On normal karyotypes,
\textit{OneD} fared equally well as state-of-the-art methods, making it a
safe choice for Hi-C normalization.  \textit{OneD} is fast and scales well
in terms of computing resources for resolutions up to 1 kbp.
%\textit{OneD} is implemented as an R package
%available at http://www.github.com/qenvio/dryhic.
\end{abstract}

@@

% **************************************************************
% Keep this command to avoid text of first page running into the
% first page footnotes
\enlargethispage{-65.1pt}
% **************************************************************

@@

\section{MATERIALS AND METHODS}

@@

%\begin{landscape}

@@

%\end{landscape}

@@

We developed the routines contained in the \texttt{dryhic} R package (available
at http://www.github.com/qenvio/dryhic) to
efficiently create sparse representations of contact matrices and further
apply \textit{vanilla}, \textit{ICE} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100 kbp, but we found no major differences at different
resolutions (not shown).

@@

\section{RESULTS}

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]{img/figure1.pdf}}
\caption{Total amount of contacts per bin. A. Non linear relationship
between the number of restriction sites and the total number of
contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. The brown line represents the raw
signal, the blue line represents the signal after bias correction, the
black line represents the signal after bias and copy number (CN) correction.
The long arm of chromosome 17 (the region corresponding to 20-80 Mbp) is
present in four copies, explaining that the signal is about twice higher
than for the short arm.}
\label{fig:totals}
\end{figure}

@@

\begin{figure}
	\centerline{\includegraphics[width=.39\textwidth]
{img/copy_number_figure2.pdf}}
\caption{Pearson correlation between the total number of contacts per bin
and an independent estimation of the copy number (COSMIC). Left panel
T47D breast cancer cell line, right panel K562 leukemia cell line.}
\label{fig:copy_number}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotypes. A and
B. Average changes compared to the raw data on the T47D and K562 sets. The
bars represent 95\% confidence intervals centered on the mean difference
of the correlation score between a given correction method and the raw
data. C and D. ROC curves on the T47D and K562 sets.  The areas under the
curve are indicated in the bottom right corner. The color code is the same
as panels A and B. The brown lines correspond to raw matrices. All results
in this figure are based on Pearson correlations between the observed over
expected counts.}
\label{fig:aberrant}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {img/correlation_normal_figure4.pdf}}
\caption{
Results of the comparison between samples with normal karyotypes. A.
Average changes compared to the raw data on the mouse data set.  B. ROC
curves on the mouse data set. The legend is as in
Figure~\ref{fig:aberrant}.}
\label{fig:normal}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {img/figure_cnv_correction.pdf}}
\caption{
Copy number correction A. Detail of a Hi-C matrix normalized with
different methods. The central portion has an increased copy number, which
affects normalization. \textit{ICE} fades it away, \textit{LGF} enhances
it and \textit{OneD} reduces the signal by about half. B. Profile of the
total amount of contacts after copy numer correction. The plot shows the
same region as panel A. The brown line represents the raw signal, the blue
line represents the signal after bias correction, the black line
represents the signal after bias and copy number (CN) correction.}
\label{fig:cnv_correction}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.49\textwidth]
  {img/figure_benchmark_time.pdf}}
\caption{
Computing time of the bias correction methods. A. Total time for the
entire genome at 100 kbp resolution. Each dot corresponds to one sample.
The only method faster than ours under performs in all sample comparisons.
B. Time to correct a reduced genome (chr19-22) of one sample at different
resolutions. Note the logarithmic scale on the y-axis on both panels.}
\label{fig:times}
\end{figure}

\section{DISCUSSION}

@@

\section{CONCLUSION}

@@

\section{ACKNOWLEDGEMENTS}

@@

\section{FUNDING}

@@

\subsubsection{Conflict of interest statement.} None declared.
\newpage
